---
title: "ECD_Supp_Mat"
output: html_document
author: Matt Barbour
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How does character displacement affect resource dynamics?

Despite the numerous models that have been developed to examine the ecological conditions resulting in character displacement, none make predictions about how character divergence influences resource abundances or stability (except perhaps Lawlor and Smith 1976). This is a bit surprising, given that explicit modeling of resource dynamics has been shown to give key insight to the process leading to character divergence.

Intuitively, you might expect character divergence to decrease resource abundances. The rationale being that consumers are evolving to be specialists that are more effective at attacking a particular resource. This however is not necessarily true. In fact, the effect of character displacement on resource dynamics depends on both the foraging scenario and trade-off in attack rates. For example, let's start with the simplest model of competition that explicitly models resource dynamics is MacArthur's model of competition (MacArthur 1972):

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-a_{11}R_1C_1-a_{21}R_1C_2$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-a_{12}R_2C_1-a_{22}R_2C_2$$
$$\frac{dC_1}{dt}=e_{11}a_{11}R_1C_1+e_{12}a_{12}R_2C_1-m_1C_1$$
$$\frac{dC_2}{dt}=e_{21}a_{21}R_1C_2+e_{22}a_{22}R_2C_2-m_2C_2$$

Now, let's examine the simplest scenario where resources are equivalent ($r=r_i$ and $K=K_i$) and consumers have equivalent conversion efficiencies ($e=e_{ii}=e_{ij}$) except that their attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). It can be shown that, at the equilibrium where both consumers and resources are present, resource abundances are equivalent ($R_i=R_1=R_2$) and are determined by the following equation:

$$\hat{R_i}=\frac{m}{e(a_{ii}+a_{aij})}$$
Note that a key determinant of resource abundance is the consumer's total attack rate, $a_{ii}+a_{ij}$. An increased in total attack rate results in a decrease in resource abundances. Now let's assume that the evolution of consumer attack rates are constrained by a linear trade-off, i.e. a unit increase in $a_{ii}$ results in a unit decrease in $a_{ij}$. Although such a trade-off results in divergent character displacement (Abrams 1986), there is no net change in resource abundances when consumer's diverge. Although this equilibrium equation for resource abundances was derived for the scenario where both consumers and both resources are present, it predicts well the abundance of resources when there is only one consumer as well. This is likely because a consumer, in the absence of competition, evolves to be a generalist that has equal attack rates on each resources, resulting in equivalent resource abundances. 

Now let's consider a different type of trade-off in consumer attack rates. We know from past work that consumers still undergo character divergence regardless of the shape of the trade-off, even though the trade-off does influence the magnitude of displacement (Lawlor and Smith 1976). We specified the following trade-off function (Sargent and Otto, 2006):

$$\Big(\frac{a_{ii}}{A}\Big)^n+\Big(\frac{a_{ij}}{A}\Big)^n=1$$
where $A$ is the total investment in attack rates and $n$ describes the shape of the trade-off. This trade-off has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$). When $n=1$, the trade-off function is linear, and all combinations of $a_ii$ and $a_ij$ have the same total attack rate.

```{r Trade-off Function, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

```


In contrast, the shape of the trade-off qualitatively affects the relationship between divergence and resource abundance. As mentioned previously, a linear trade-off results in no change in resource densities. If the trade-off is convex, then resource densities can actually increase under character displacement! This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{aij}$) and decreases as consumers diverge. Note though that detecting this effect in nature may be difficult, since a convex trade-off results in relatively small displacement relative to other trade-off forms (Lawlor and Smith 1976). When the trade-off is concave, however, we see that character divergence suppresses resource densities.

Most models of character displacement have assumed a scenario where consumers can forage for both resources simultaneously (i.e. resources occur in the same microhabitat). This assumption may be valid for some foraging scenarios, e.g. Darwin's finches foraging for seeds. However, in many situations, consumers forage for resources that occur in distinct microhabitats (McCann et al. 2005). The only chracter displacement model that we are aware of that modeled resources in distinct microhabitat was one examined by Lawlor and Smith (1976). They found that consumers still underwent character divergence, regardless of the foraging scenario. Interestingly, we find that the foraging scenario qualitatively affects the relationship between character divergence and resource abundances. Specifically, if consumers are competing for resources that occur in distinct microhabitats, then consumer divergence always results in resource suppression. This occurs because now the consumers effective attack rate is influenced by its habitat preference. Note also that this resource suppression occurs regardless of the shape of the trade-off, although the shape of the trade-off can amplify the effect of character divergence on resource supression. The type of trade-off does affect the amount of character displacement though, and actually, we see again the a concave trade-off results in less displacement and less of an effect on resource abundances; whereas a convex trade-off results in the largest displacement and the largest effect on resource abundances. This effect of spatial context on the relationship between character displacement and resource abundances appears to be quite general. For example, in the appendix, we analyze a model that was designed to capture the spatial coupling of consumers in many food-web context (McCann 2005). Here, we again find that divergence results in a negative effect on resource abundances. 

We would like to note that the scenario where the relationship between character divergence and resource suppression is the strongest is when the trade-off in attack rates is concave and consumers are competiting for resources that occupy distinct habitat types. Given that prior work in threespine stickleback has shown a concave trade-off in feeding rates (Schluter 1993; Arnegard et al. 2014) and that stickleback ecotypes are competing for resources that occur in distinct microhabitats (benthic vs. limnetic), we predict that character divergence in sticklebacks would supress resource densities. 

This relationship between character divergence and resource abundances begs the question: how does character divergence affect food-web stability? This question has been addressed empirically by Pimentel as well as with a conceptual model. Lawlor and Smith though did a detailed theoretical analysis of how character divergence affects three different aspects of stability. They looked at stability through the lens of competition. The first aspect was whether character divergence resulted in mutual invasibility of consumers, a common metric of "stable coexistence" in competition models. Another very interesting aspect they looked at was a global metric that examined the persistence of the system to perturbations in resources. Here, they conclude the divergence increases community stability, because it expands the range of resource perturbations that a system can withstand before a consumer going extinct. Note though, that they overlooked that there conclusions do not actually quantify the effects of ecological character displacement on stability because they do not compare scenarios where a consumer is in allopatry and one where consumers are in sympatry. Finally, they also look at the Lyapunov exponent (local stability), and while they find a negative effect of divergence, they suggest that it would likely be overshadowed by the stabilizing effects of divergence on mutual invasibility and global stability.

Here we focus on local stability, which makes the most sense in a food-web context. We find that divergence follows the typical "check-mark" stability pattern that has been described recently in theoretical food-web models.

Importantly, this metric of stability allows us to compare the stability of three-species vs. four-species system, which is key for understanding how ecological character displacement affects food-web dynamics.


## Testing the generality of our predictions

So far, our theoretical predictions of the effects of consumer divergence on resource dynamics has considered a simplistic scenario, where resources are equivalent and consumers have equivalent conversion efficiences, but have mirror image attack rates on each resource. We conjecture that these results are quite general, given that the process of character displacement is quite general. To explore the generality of our predictions, we examined whether our conclusions still held under a foraging scenario that was designed to mimic the stickleback lake food web. To do this, we used empirical relationships between body size and the parameters that govern consumer-resource dynamics to parameterize an eco-evolutionary model. Specifically, we used consumer and resource biomasses 
